The Proofs

### 2023-11-13 proofs/tex/equiv-lemmas.tex
wip
Restructure proofs

\begin{lemma} [Declarative Equivalence is invariant under bijections]
  \label{lemma:decl-equiv-bijection}
  Suppose $[[μ]]$ is a bijection $[[μ : varset1 ↔ varset2]]$, then
  \begin{itemize}
    \item[$+$] $[[iP1 ≈ iP2]]$ implies $[[ [μ]iP1 ≈ [μ]iP2 ]]$,
      and there exists an inference tree of $[[ [μ]iP1 ≈ [μ]iP2 ]]$
      with the same shape as the one inferring $[[iP1 ≈ iP2]]$;
    \item[$-$] $[[iN1 ≈ iN2]]$ implies $[[ [μ]iN1 ≈ [μ]iN2 ]]$, 
      and there exists an inference tree of $[[ [μ]iN1 ≈ [μ]iN2 ]]$
      with the same shape as the one inferring $[[iN1 ≈ iN2]]$.
  \end{itemize}
\end{lemma}
\begin{proof}
  We prove it by induction on $[[iP1 ≈ iP2]]$ and mutually, on $[[iN1 ≈ iN2]]$.
  Let us consider the last rule used in the derivation.
  \begin{caseof}
    \item \ruleref{\ottdruleEOneForallLabel}\\
      Then we decompose $[[iN1]]$ as $[[∀pas1.iM1]]$ and $[[iN2]]$ as $[[∀pas2.iM2]]$,
      where $[[iM1]]$ and $[[iM2]]$ do not start with $[[∀]]$-quantifiers.
      where $[[|pas1| + |pas2| > 0]]$.
      By convention, let us assume that
      $[[pas1]]$ and $[[pas2]]$ are disjoint form 
      $[[varset2]]$ and $[[varset1]]$.

@@

\begin{lemma}[Declarative equivalence is transitive]
  \hfill
  \label{lemma:decl-equiv-transitivity}
  \begin{itemize}
    \item[$+$] if $[[iP1 ≈ iP2]]$ and $[[iP2 ≈ iP3]]$ then $[[iP1 ≈ iP3]]$,
    \item[$-$] if $[[iN1 ≈ iN2]]$ and $[[iN2 ≈ iN3]]$ then $[[iN1 ≈ iN3]]$.
  \end{itemize}
\end{lemma}
\begin{proof}
  We prove it by
  $\size{[[iP1 ≈ iP2]]} + \size{[[iP2 ≈ iP3]]}$
  and mutually, $\size{[[iN1 ≈ iN2]]} + \size{[[iN2 ≈ iN3]]}$, where 
  by size, we mean the size of the nodes in the corresponding inference tree.

@@

\begin{lemma}[Type well-formedness is invariant under equivalence]
  \label{lemma:wf-equiv}
  Mutual subtyping implies declarative equivalence.
  \begin{itemize}
  \item[$+$] if $[[iP ≈ iQ]]$ then $[[Γ ⊢ iP]] \iff [[Γ ⊢ iQ]]$,
  \item[$-$] if $[[iN ≈ iM]]$ then $[[Γ ⊢ iN]] \iff [[Γ ⊢ iM]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  We prove it by induction on $[[iP ≈ iQ]]$ and mutually, on $[[iN ≈ iM]]$.
  Let us consider the last rule used in the derivation.
  \begin{caseof}
    \item \ruleref{\ottdruleEOneNVarLabel}, 
      that is $[[iN ≈ iM]]$ has shape $[[α⁻ ≈ α⁻]]$.\\
      Than $[[Γ ⊢ iP]] \iff [[Γ ⊢ iQ]]$ is rewritten as $[[Γ ⊢ α⁻]] \iff [[Γ ⊢ α⁻]]$,
      which holds trivially.
    \item \ruleref{\ottdruleEOneShiftULabel},
      that is $[[iN ≈ iM]]$ has shape $[[↑iP ≈ ↑iQ]]$.\\
      By inversion, $[[iP ≈ iQ]]$, and by the induction hypothesis,
      $[[Γ ⊢ iP]] \iff [[Γ ⊢ iQ]]$. 
      Also notice that $[[Γ ⊢ ↑iP]] \iff [[Γ ⊢ iP]]$ 
      and $[[Γ ⊢ ↑iQ]] \iff [[Γ ⊢ iQ]]$ by inversion and \ruleref{\ottdruleWFTShiftULabel}.
      This way, $[[Γ ⊢ ↑iP]] \iff [[Γ ⊢ iP]] \iff [[Γ ⊢ iQ]] \iff [[Γ ⊢ ↑iQ]]$.
    \item \ruleref{\ottdruleEOneArrowLabel},
      that is $[[iN ≈ iM]]$ has shape $[[iP → iN' ≈ iQ → iM']]$.\\
      Then by inversion, $[[iP ≈ iQ]]$ and $[[iN' ≈ iM']]$, 
      and by the induction hypothesis, $[[Γ ⊢ iP]] \iff [[Γ ⊢ iQ]]$ and $[[Γ ⊢ iN']] \iff [[Γ ⊢ iM']]$.
      $
      \begin{aligned}[t]
        [[Γ ⊢ iP → iN']]  &\iff [[Γ ⊢ iP]] \text{ and } [[Γ ⊢ iN']] 
                          && \text{by inversion and \ruleref{\ottdruleWFTArrowLabel}}\\
                          &\iff [[Γ ⊢ iQ]] \text{ and } [[Γ ⊢ iM']]
                          && \text{as noted above}\\
                          &\iff [[Γ ⊢ iQ → iM']]
                          && \text{by \ruleref{\ottdruleWFTArrowLabel} and inversion}\\
      \end{aligned}
      $\\
    \item \ruleref{\ottdruleEOneForallLabel},
      that is $[[iN ≈ iM]]$ has shape $[[∀pas.iN' ≈ ∀pbs.iM']]$.\\
      By inversion, $[[∀pas.iN' ≈ ∀pbs.iM']]$ means that
      $[[{pas} ∩ fv iM = ∅]]$ and 
      that there exists a bijection on variables 
      $[[mu : ({pbs} ∩ fv iM') ↔ ({pas} ∩ fv iN')]]$
      such that $[[iN' ≈ [mu] iM']]$.

@@

\begin{corollary}[Normalization preserves well-formedness]
  \label{corollary:wf-nf}
  \hfill
  \begin{itemize}
  \item[$+$] $[[Γ ⊢ iP]] \iff [[Γ ⊢ nf(iP)]]$,
  \item[$-$] $[[Γ ⊢ iN]] \iff [[Γ ⊢ nf(iN)]]$
  \end{itemize}
\end{corollary}
\begin{proof}
  Immediately from \cref{lemma:wf-equiv,lemma:normalization-soundness}.
\end{proof}

\begin{corollary}[Normalization preserves well-formedness of substitution]
  \label{corollary:wf-s-nf}
  \hfill \\
   $[[Γ2 ⊢ σ : Γ1]] \iff [[Γ2 ⊢ nf(σ) : Γ1]]$
\end{corollary}
\begin{proof}
  Let us prove the forward direction.
  Suppose that $[[α± ∊ Γ1]]$.  Let us show that $[[Γ2 ⊢ [nf(σ)]α±]]$.
  By the definition of substitution normalization,
  $[[ [nf(σ)]α± = nf([σ]α±) ]]$. Then by \cref{corollary:wf-nf},
  to show $[[Γ2 ⊢ nf([σ]α±)]]$, it suffices to show $[[Γ2 ⊢ [σ]α±]]$,
  which holds by the assumption $[[Γ2 ⊢ σ : Γ1]]$.

  The backward direction is proved analogously.
\end{proof}

\begin{lemma}[Normalization preserves substitution signature]
  \label{lemma:norm-subst-sig}
  Suppose that $[[σ]]$ is a substitution, $[[Γ1]]$ and $[[Γ2]]$ are contexts.
  Then $[[Γ2 ⊢ σ : Γ1]]$ implies $[[Γ2 ⊢ nf(σ) : Γ1]]$.
\end{lemma}
\begin{proof}
  Suppose that $[[α± ∊ Γ1]]$. 
  Then by \cref{corollary:wf-nf}, $[[Γ2 ⊢ nf([σ]α±)]] = [[ [nf(σ)]α± ]]$ 
  is equivalent to $[[Γ2 ⊢ [σ]α±]]$.

  Suppose that $[[α± ∉ Γ1]]$. 
  $[[Γ2 ⊢ σ : Γ1]]$ means that $[[ [σ]α± = α± ]]$, 
  and then $[[ [nf(σ)]α± ]] = [[nf([σ]α±)]] = [[nf(α±)]] = [[α±]]$.
\end{proof}

\begin{lemma}[Soundness of equivalence]
  \label{lemma:equiv-soundness}
  Declarative equivalence implies mutual subtyping.
  \begin{itemize}
    \item[$+$] if $[[Γ ⊢ iP]]$, $[[Γ ⊢ iQ]]$, and $[[iP ≈ iQ]]$ then $[[Γ ⊢ iP ≈ iQ]]$,
    \item[$-$] if $[[Γ ⊢ iN]]$, $[[Γ ⊢ iM]]$, and $[[iN ≈ iM]]$ then $[[Γ ⊢ iN ≈ iM]]$.
  \end{itemize}
\end{lemma}
\begin{proof}
  We prove it by mutual induction on $[[iP ≈ iQ]]$ and $[[iN ≈ iM]]$.
  \begin{caseof}
    \item $[[a⁻ ≈ a⁻]]$\\
      Then $[[Γ ⊢ a⁻ ≤ a⁻]]$ by \ruleref{\ottdruleDOneNVarLabel},
      which immediately implies $[[Γ ⊢ a⁻ ≈ a⁻]]$ by \ruleref{\ottdruleDOneNDefLabel}.

@@

\begin{corollary}[Normalization is sound w.r.t. subtyping-induced equivalence] 
  \label{corollary:nf-sound-wrt-subt-equiv}
  \hfill
  \begin{itemize}
    \item [$+$] if $[[Γ ⊢ iP]]$ then $[[Γ ⊢ iP ≈ nf(iP)]]$,
    \item [$-$] if $[[Γ ⊢ iN]]$ then $[[Γ ⊢ iN ≈ nf(iN)]]$.
  \end{itemize}
\end{corollary}
\begin{proof}
  Immediately from \cref{lemma:normalization-soundness,corollary:wf-nf,lemma:equiv-soundness}.
\end{proof}

\begin{corollary}[Normalization preserves subtyping] 
  \label{corollary:nf-pres-subt}
  Assuming all the types are well-formed in context $[[Γ]]$,
  \begin{itemize}
    \item [$+$] $[[Γ ⊢ iP ≥ iQ]] \iff [[Γ ⊢ nf(iP) ≥ nf(iQ)]]$,
    \item [$-$] $[[Γ ⊢ iN ≤ iM]] \iff [[Γ ⊢ nf(iN) ≤ nf(iM)]]$.
  \end{itemize}
\end{corollary}
\begin{proof}
  \hfill
  \begin{itemize}
    \item [$+$]  
    \begin{itemize}
      \item [$\Rightarrow$] Let us assume $[[Γ ⊢ iP ≥ iQ]]$.
        By \cref{corollary:nf-sound-wrt-subt-equiv},
        $[[Γ ⊢ iP ≈ nf(iP)]]$ and $[[Γ ⊢ iQ ≈ nf(iQ)]]$, 
        in particular, by inversion, 
        $[[Γ ⊢ nf(iP) ≥ iP]]$ and $[[Γ ⊢ iQ ≥ nf(iQ)]]$.
        Then by transitivity of subtyping 
        (\cref{lemma:subtyping-transitivity}), 
        $[[Γ ⊢ nf(iP) ≥ nf(iQ)]]$.
      \item [$\Leftarrow$] Let us assume $[[Γ ⊢ nf(iP) ≥ nf(iQ)]]$.
        Also by \cref{corollary:nf-sound-wrt-subt-equiv}
        and inversion, 
        $[[Γ ⊢ iP ≥ nf(iP)]]$ and $[[Γ ⊢ nf(iQ) ≥ iQ]]$.
        Then by the transitivity, $[[Γ ⊢ iP ≥ iQ]]$.
    \end{itemize}
    \item [$-$] The negative case is proved symmetrically.
  \end{itemize}
\end{proof}

\begin{lemma}[Subtyping induced by disjoint substitutions]
  \label{lemma:subt-ind-disj-subst}
  If two disjoint substitutions induce subtyping, they are degenerate (so is the
  subtyping).
  Suppose that $[[Γ ⊢ σ1 : Γ1]]$ and $[[Γ ⊢ σ2 : Γ1]]$,
  where $[[Γi ⊆ Γ]]$ and $[[Γ1 ∩ Γ2= ∅]]$. Then
  \begin{itemize}
  \item[$-$] assuming $[[Γ ⊢ iN]]$,~
    $[[Γ ⊢ [σ1]iN ≤ [σ2]iN]]$ implies $[[Γ ⊢ σi ≈ id : fv iN]]$
  \item[$+$] assuming $[[Γ ⊢ iP]]$,~
    $[[Γ ⊢ [σ1]iP ≥ [σ2]iP]]$ implies $[[Γ ⊢ σi ≈ id : fv iP]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  Proof by induciton on $[[Γ ⊢ iN]]$ (and mutually on $[[Γ ⊢ iP]]$).
  \begin{caseof}
    \item $[[iN]] = [[α⁻]]$\\
      Then $[[Γ ⊢ [σ1]iN ≤ [σ2]iN]]$ is rewritten as $[[Γ ⊢ [σ1]α⁻ ≤ [σ2]α⁻]]$.
      Let us consider the following cases:
      \begin{caseof}
      \item $[[α⁻ ∉ Γ1]]$ and $[[α⁻ ∉ Γ2]]$ \label{case:var-not-in-ctxts}\\
        Then $[[Γ ⊢ σi ≈ id : {α⁻}]]$ holds immediately,
        since $[[ [σi] α⁻]] = [[ [id] α⁻]] = [[α⁻]]$ and
        $[[Γ ⊢ α⁻ ≈ α⁻]]$.
      \item $[[α⁻ ∊ Γ1]]$ and $[[α⁻ ∊ Γ2]]$\\
        This case is not possible by assumption: $[[Γ1 ∩ Γ2= ∅]]$.
      \item $[[α⁻ ∊ Γ1]]$ and $[[α⁻ ∉ Γ2]]$\\
        Then we have $[[Γ ⊢ [σ1]α⁻ ≤ α⁻]]$,
        which by \cref{corollary:vars-no-proper-subtypes} means $[[Γ ⊢ [σ1]α⁻ ≈ α⁻]]$,
        and hence, $[[Γ ⊢ σ1 ≈ id : {α⁻}]]$.

@@

\begin{corollary}[Substitution cannot induce proper subtypes or supertypes] \label{corollary:subst-proper-subt}
  Assuming all mentioned types are well-formed in $[[Γ]]$ and $[[σ]]$ is a
  substitution $[[Γ ⊢ σ : Γ]]$,
  \begin{align*}
    [[Γ ⊢ [σ]iN ≤ iN]] ~&\Rightarrow~ [[Γ ⊢ [σ]iN ≈ iN]]
                          \text{ and } [[Γ ⊢ σ ≈ id :  fv iN]] \\
    [[Γ ⊢ iN ≤ [σ]iN]] ~&\Rightarrow~ [[Γ ⊢ iN ≈ [σ]iN]]
                          \text{ and } [[Γ ⊢ σ ≈ id :  fv iN]] \\
    [[Γ ⊢ [σ]iP ≥ iP]] ~&\Rightarrow~ [[Γ ⊢ [σ]iP ≈ iP]]
                          \text{ and } [[Γ ⊢ σ ≈ id :  fv iP]] \\
    [[Γ ⊢ iP ≥ [σ]iP]] ~&\Rightarrow~ [[Γ ⊢ iP ≈ [σ]iP]]
                          \text{ and } [[Γ ⊢ σ ≈ id :  fv iP]] \\
  \end{align*}
\end{corollary}

\begin{lemma} \label{lemma:mutual-subst-subtyping}
  Assuming that the mentioned types ($[[iP]]$, $[[iQ]]$, $[[iN]]$, and $[[iM]]$)
  are well-formed in $[[Γ]]$ and that the substitutions ($[[σ1]]$ and $[[σ2]]$) have signature $[[Γ ⊢ σi : Γ]]$,
  \begin{itemize}
  \item[$+$] if $[[Γ ⊢ [σ1] iP ≥ iQ]]$ and $[[Γ ⊢ [σ2] iQ ≥ iP]]$\\
    then there exists a bijection $[[μ : fv iP ↔ fv iQ]]$ such that
    $[[Γ ⊢ σ1 ≈ Sub μ :  fv iP]]$ and $[[Γ ⊢ σ2 ≈ Sub μ-1 :  fv iQ]]$;
  \item[$-$] if $[[Γ ⊢ [σ1] iN ≤ iM]]$ and $[[Γ ⊢ [σ2] iN ≤ iM]]$\\
    then there exists a bijection $[[μ : fv iN ↔ fv iM]]$ such that
    $[[Γ ⊢ σ1 ≈ Sub μ :  fv iN]]$ and $[[Γ ⊢ σ2 ≈ Sub μ-1 :  fv iM]]$.
  \end{itemize}
\end{lemma}
\begin{proof}
  \hfill
  \begin{itemize}
  \item[$+$]
    Applying $[[σ2]]$ to both sides of
    $[[Γ ⊢ [σ1] iP ≥ iQ]]$ (by \cref{lemma:subst-pres-subt}),
    we have: $[[Γ ⊢ [σ2 ○ σ1] iP ≥ [σ2]iQ]]$.
    Composing it with $[[Γ ⊢ [σ2] iQ ≥ iP]]$ by transitivity 
    (\cref{lemma:subtyping-transitivity}),
    we have $[[Γ ⊢ [σ2 ○ σ1] iP ≥ iP]]$.
    Then by \cref{corollary:subst-proper-subt},
    $[[Γ ⊢ σ2 ○ σ1 ≈ id :  fv iP]]$.
    By a symmetric argument, we also have:
    $[[Γ ⊢ σ1 ○ σ2 ≈ id :  fv iQ]]$.

@@

\begin{lemma}[Equivalent substitution act equivalently]
  \label{lemma:equiv-subst-on-same-term}
  Suppose that $[[Γ' ⊢ σ1 : Γ]]$ and $[[Γ' ⊢ σ2 : Γ]]$
  are substitutions equivalent on their domain, that is $[[Γ' ⊢ σ1 ≈ σ2 : Γ]]$.
  Then
  \begin{itemize}
    \item[$+$] for any $[[Γ ⊢ iP]]$, $[[Γ' ⊢ [σ1]iP ≈ [σ2]iP]]$;
    \item[$-$] for any $[[Γ ⊢ iN]]$, $[[Γ' ⊢ [σ1]iN ≈ [σ2]iN]]$.
  \end{itemize}
\end{lemma}
\begin{proof}
  We prove it by induction on $[[iP]]$ (and mutually on $[[iN]]$).
  \begin{caseof}
    \item $[[iN = α⁻]]$\\
      Then since by inversion, $[[α⁻ ∊ Γ]]$, 
      $[[Γ' ⊢ [σ1]α⁻ ≈ [σ2]α⁻]]$ holds by definition of $[[Γ' ⊢ σ1 ≈ σ2 : Γ]]$.
    \item $[[iN = ↑iP]]$\\
      Then by inversion, $[[Γ ⊢ iP]]$.
      By the induction hypothesis, $[[Γ' ⊢ [σ1]iP ≈ [σ2]iP]]$, 
      Then by \ruleref{\ottdruleDOneShiftULabel}, 
      $[[Γ' ⊢ ↑[σ1]iP ≤ ↑[σ2]iP]]$, and symmetrically, $[[Γ' ⊢ ↑[σ2]iP ≤ ↑[σ1]iP]]$,
      together meaning that $[[Γ' ⊢ ↑[σ1]iP ≈ ↑[σ2]iP]]$,
      or equivalently, $[[Γ' ⊢ [σ1]↑iP ≈ [σ2]↑iP]]$.
    \item $[[iN = iP → iM]]$\\
      Then by inversion, $[[Γ ⊢ iP]]$ and $[[Γ ⊢ iM]]$.
      By the induction hypothesis, $[[Γ' ⊢ [σ1]iP ≈ [σ2]iP]]$ and
      $[[Γ' ⊢ [σ1]iM ≈ [σ2]iM]]$,
      that is $[[Γ' ⊢ [σ1]iP ≥ [σ2]iP]]$, $[[Γ' ⊢ [σ2]iP ≥ [σ1]iP]]$,
      $[[Γ' ⊢ [σ1]iM ≤ [σ2]iM]]$, and $[[Γ' ⊢ [σ2]iM ≤ [σ1]iM]]$.
      Then by \ruleref{\ottdruleDOneArrowLabel}, $[[Γ' ⊢ [σ1]iP → [σ1]iM ≤ [σ2]iP → [σ2]iM]]$, 
      and again by \ruleref{\ottdruleDOneArrowLabel}, $[[Γ' ⊢ [σ2]iP → [σ2]iM ≤ [σ1]iP → [σ1]iM]]$.
      This way, $[[Γ' ⊢ [σ1]iP → [σ1]iM ≈ [σ2]iP → [σ2]iM]]$,
      or equivalently, $[[Γ' ⊢ [σ1](iP → iM) ≈ [σ2](iP → iM)]]$.
    \item $[[iN = ∀pas.iM]]$
      We can assume that $[[pas]]$ is disjoint from $[[Γ]]$ and $[[Γ']]$.
      By inversion, $[[Γ ⊢ ∀pas.iM]]$ implies $[[Γ, pas ⊢ iM]]$.
      Notice that $[[Γ' ⊢ σi : Γ]]$ and 
      $[[Γ' ⊢ σ1 ≈ σ2 : Γ]]$
      can be extended to 
      $[[Γ', pas ⊢ σi : Γ, pas]]$
      and 
      $[[Γ', pas ⊢ σ1 ≈ σ2 : Γ, pas]]$
      by \cref{lemma:subst-domain-weakening}.
      Then by the induction hypothesis,
      $[[Γ', pas ⊢ [σ1]iM ≈ [σ2]iM]]$,
      meaning by inversion that
      $[[Γ', pas ⊢ [σ1]iM ≤ [σ2]iM]]$ and 
      $[[Γ', pas ⊢ [σ2]iM ≤ [σ1]iM]]$.

@@

\begin{corollary}[Normalization is complete w.r.t. subtyping-induced equivalence]
  \label{corollary:nf-complete-wrt-subt-equiv}
  Assuming all the types below are well-formed in $[[Γ]]$:
  \begin{itemize}
    \item [$+$] if $[[Γ ⊢ iP ≈ iQ]]$ then $[[nf(iP) = nf(iQ)]]$,
    \item [$-$] if $[[Γ ⊢ iN ≈ iM]]$ then $[[nf(iN) = nf(iM)]]$.
  \end{itemize}
\end{corollary}  
\begin{proof}
  Immediately from \cref{lemma:equiv-completeness,lemma:normalization-completeness}.
\end{proof}

>>>>

\lemmaDeclEquivBijection*
\begin{proof}
  We prove it by induction on $[[iP1 ≈ iP2]]$ and mutually, on $[[iN1 ≈ iN2]]$.
  Let us consider the last rule used in the derivation.
  \begin{caseof}
    \item \ruleref{\ottdruleEOneForallLabel}\\
      Then we decompose $[[iN1]]$ as $[[∀pas1.iM1]]$ and $[[iN2]]$ as $[[∀pas2.iM2]]$,
      where $[[iM1]]$ and $[[iM2]]$ do not start with $[[∀]]$-quantifiers.
      where $[[|pas1| + |pas2| > 0]]$.
      By convention, let us assume that
      $[[pas1]]$ and $[[pas2]]$ are disjoint form 
      $[[varset2]]$ and $[[varset1]]$.

@@

\lemmaEquivFV*
\begin{proof}
  Mutual induction on $[[iN ≈ iM]]$ and $[[iP ≈ iQ]]$
  The base cases 
  (\ruleref{\ottdruleEOneNVarLabel} and \ruleref{\ottdruleEOnePVarLabel})
  are trivial.
  So are \ruleref{\ottdruleEOneShiftULabel}, 
  \ruleref{\ottdruleEOneShiftDLabel}, and
  \ruleref{\ottdruleEOneArrowLabel}, where
  the required property follows from the induction hypothesis.

  Let us consider the case when the equivalence is formed by
  \ruleref{\ottdruleEOneForallLabel}, 
  that is the equivalence has a shape 
  $[[∀pas.iN ≈ ∀pbs.iM]]$, 
  and by inversion,
  there is a bijection 
  $[[mu : ({pbs} ∩ fv iM) ↔ ({pas} ∩ fv iN)]]$
  such that $[[iN ≈ [mu] iM]]$, which 
  by the induction hypothesis means
  $[[ fv iN = fv [mu]iM ]] = [[ [mu] fv iM]] $.

  Let us ensure by alpha-equivalence that 
  $[[pas]]$ is disjoint from $[[fv iM]]$.
  Then $[[ (fv ∀pbs.iM) \ {pas}]] = [[fv ∀pbs.iM]]$.
  Then we apply the following chain of equalities:
  $[[fv ∀pas.iN = fv iN \ {pas}]] = 
  [[ ([mu] fv iM) \ {pas}]] = 
  [[ [mu] (fv ∀pbs.iM ∪ ({pbs} ∩ fv iM)) \ {pas}]] = 
  [[ ([mu] fv ∀pbs.iM ∪ [mu]({pbs} ∩ fv iM)) \ {pas}]] = 
  [[ ([mu] fv ∀pbs.iM) \ {pas}]] = 
  [[ (fv ∀pbs.iM) \ {pas}]] = 
  [[  fv ∀pbs.iM ]]$.

  Symmetrically, we prove the case when the equivalence is formed by
  \ruleref{\ottdruleEOneExistsLabel}.
\end{proof}

\lemmaDeclEquivTransitivity*
\begin{proof}
  We prove it by
  $\size{[[iP1 ≈ iP2]]} + \size{[[iP2 ≈ iP3]]}$
  and mutually, $\size{[[iN1 ≈ iN2]]} + \size{[[iN2 ≈ iN3]]}$, where 
  by size, we mean the size of the nodes in the corresponding inference tree.

@@

\lemmaWfEquiv*
\begin{proof}
  We prove it by induction on $[[iP ≈ iQ]]$ and mutually, on $[[iN ≈ iM]]$.
  Let us consider the last rule used in the derivation.
  \begin{caseof}
    \item \ruleref{\ottdruleEOneNVarLabel}, 
      that is $[[iN ≈ iM]]$ has shape $[[α⁻ ≈ α⁻]]$.\\
      Than $[[Γ ⊢ iP]] \iff [[Γ ⊢ iQ]]$ is rewritten as $[[Γ ⊢ α⁻]] \iff [[Γ ⊢ α⁻]]$,
      which holds trivially.
    \item \ruleref{\ottdruleEOneShiftULabel},
      that is $[[iN ≈ iM]]$ has shape $[[↑iP ≈ ↑iQ]]$.\\
      By inversion, $[[iP ≈ iQ]]$, and by the induction hypothesis,
      $[[Γ ⊢ iP]] \iff [[Γ ⊢ iQ]]$. 
      Also notice that $[[Γ ⊢ ↑iP]] \iff [[Γ ⊢ iP]]$ 
      and $[[Γ ⊢ ↑iQ]] \iff [[Γ ⊢ iQ]]$ by inversion and \ruleref{\ottdruleWFTShiftULabel}.
      This way, $[[Γ ⊢ ↑iP]] \iff [[Γ ⊢ iP]] \iff [[Γ ⊢ iQ]] \iff [[Γ ⊢ ↑iQ]]$.
    \item \ruleref{\ottdruleEOneArrowLabel},
      that is $[[iN ≈ iM]]$ has shape $[[iP → iN' ≈ iQ → iM']]$.\\
      Then by inversion, $[[iP ≈ iQ]]$ and $[[iN' ≈ iM']]$, 
      and by the induction hypothesis, $[[Γ ⊢ iP]] \iff [[Γ ⊢ iQ]]$ and $[[Γ ⊢ iN']] \iff [[Γ ⊢ iM']]$.

      $
      \begin{aligned}[t]
        [[Γ ⊢ iP → iN']]  &\iff [[Γ ⊢ iP]] \text{ and } [[Γ ⊢ iN']] 
                          && \text{by inversion and \ruleref{\ottdruleWFTArrowLabel}}\\
                          &\iff [[Γ ⊢ iQ]] \text{ and } [[Γ ⊢ iM']]
                          && \text{as noted above}\\
                          &\iff [[Γ ⊢ iQ → iM']]
                          && \text{by \ruleref{\ottdruleWFTArrowLabel} and inversion}\\
      \end{aligned}
      $\\
    \item \ruleref{\ottdruleEOneForallLabel},
      that is $[[iN ≈ iM]]$ has shape $[[∀pas.iN' ≈ ∀pbs.iM']]$.\\
      By inversion, $[[∀pas.iN' ≈ ∀pbs.iM']]$ means that
      $[[{pas} ∩ fv iM = ∅]]$ and 
      that there exists a bijection on variables 
      $[[mu : ({pbs} ∩ fv iM') ↔ ({pas} ∩ fv iN')]]$
      such that $[[iN' ≈ [mu] iM']]$.

@@

\lemmaEquivSoundness*
\begin{proof}
  We prove it by mutual induction on $[[iP ≈ iQ]]$ and $[[iN ≈ iM]]$.
  \begin{caseof}
    \item $[[a⁻ ≈ a⁻]]$\\
      Then $[[Γ ⊢ a⁻ ≤ a⁻]]$ by \ruleref{\ottdruleDOneNVarLabel},
      which immediately implies $[[Γ ⊢ a⁻ ≈ a⁻]]$ by \ruleref{\ottdruleDOneNDefLabel}.

@@

\lemmaSubtIndDisjSubst*
\begin{proof}
  Proof by induciton on $[[Γ ⊢ iN]]$ (and mutually on $[[Γ ⊢ iP]]$).
  \begin{caseof}
    \item $[[iN]] = [[α⁻]]$\\
      Then $[[Γ ⊢ [σ1]iN ≤ [σ2]iN]]$ is rewritten as $[[Γ ⊢ [σ1]α⁻ ≤ [σ2]α⁻]]$.
      Let us consider the following cases:
      \begin{caseof}
      \item $[[α⁻ ∉ Γ1]]$ and $[[α⁻ ∉ Γ2]]$ \label{case:var-not-in-ctxts}\\
        Then $[[Γ ⊢ σi ≈ id : {α⁻}]]$ holds immediately,
        since $[[ [σi] α⁻]] = [[ [id] α⁻]] = [[α⁻]]$ and
        $[[Γ ⊢ α⁻ ≈ α⁻]]$.
      \item $[[α⁻ ∊ Γ1]]$ and $[[α⁻ ∊ Γ2]]$\\
        This case is not possible by assumption: $[[Γ1 ∩ Γ2= ∅]]$.
      \item $[[α⁻ ∊ Γ1]]$ and $[[α⁻ ∉ Γ2]]$\\
        Then we have $[[Γ ⊢ [σ1]α⁻ ≤ α⁻]]$,
        which by \cref{corollary:vars-no-proper-subtypes} means $[[Γ ⊢ [σ1]α⁻ ≈ α⁻]]$,
        and hence, $[[Γ ⊢ σ1 ≈ id : {α⁻}]]$.

@@

\corollarySubstProperSubt*

\lemmaMutualSubstSubtyping*
\begin{proof}
  \hfill
  \begin{itemize}
  \item[$+$]
    Applying $[[σ2]]$ to both sides of
    $[[Γ ⊢ [σ1] iP ≥ iQ]]$ (by \cref{lemma:subst-pres-subt}),
    we have: $[[Γ ⊢ [σ2 ○ σ1] iP ≥ [σ2]iQ]]$.
    Composing it with $[[Γ ⊢ [σ2] iQ ≥ iP]]$ by transitivity 
    (\cref{lemma:subtyping-transitivity}),
    we have $[[Γ ⊢ [σ2 ○ σ1] iP ≥ iP]]$.
    Then by \cref{corollary:subst-proper-subt},
    $[[Γ ⊢ σ2 ○ σ1 ≈ id :  fv iP]]$.
    By a symmetric argument, we also have:
    $[[Γ ⊢ σ1 ○ σ2 ≈ id :  fv iQ]]$.

@@

\lemmaEquivSubstOnSameTerm*
\begin{proof}
  We prove it by induction on $[[iP]]$ (and mutually on $[[iN]]$).
  \begin{caseof}
    \item $[[iN = α⁻]]$\\
      Then since by inversion, $[[α⁻ ∊ Γ]]$, 
      $[[Γ' ⊢ [σ1]α⁻ ≈ [σ2]α⁻]]$ holds by definition of $[[Γ' ⊢ σ1 ≈ σ2 : Γ]]$.
    \item $[[iN = ↑iP]]$\\
      Then by inversion, $[[Γ ⊢ iP]]$.
      By the induction hypothesis, $[[Γ' ⊢ [σ1]iP ≈ [σ2]iP]]$, 
      Then by \ruleref{\ottdruleDOneShiftULabel}, 
      $[[Γ' ⊢ ↑[σ1]iP ≤ ↑[σ2]iP]]$, and symmetrically, $[[Γ' ⊢ ↑[σ2]iP ≤ ↑[σ1]iP]]$,
      together meaning that $[[Γ' ⊢ ↑[σ1]iP ≈ ↑[σ2]iP]]$,
      or equivalently, $[[Γ' ⊢ [σ1]↑iP ≈ [σ2]↑iP]]$.
    \item $[[iN = iP → iM]]$\\
      Then by inversion, $[[Γ ⊢ iP]]$ and $[[Γ ⊢ iM]]$.
      By the induction hypothesis, $[[Γ' ⊢ [σ1]iP ≈ [σ2]iP]]$ and
      $[[Γ' ⊢ [σ1]iM ≈ [σ2]iM]]$,
      that is $[[Γ' ⊢ [σ1]iP ≥ [σ2]iP]]$, $[[Γ' ⊢ [σ2]iP ≥ [σ1]iP]]$,
      $[[Γ' ⊢ [σ1]iM ≤ [σ2]iM]]$, and $[[Γ' ⊢ [σ2]iM ≤ [σ1]iM]]$.
      Then by \ruleref{\ottdruleDOneArrowLabel}, $[[Γ' ⊢ [σ1]iP → [σ1]iM ≤ [σ2]iP → [σ2]iM]]$, 
      and again by \ruleref{\ottdruleDOneArrowLabel}, $[[Γ' ⊢ [σ2]iP → [σ2]iM ≤ [σ1]iP → [σ1]iM]]$.
      This way, $[[Γ' ⊢ [σ1]iP → [σ1]iM ≈ [σ2]iP → [σ2]iM]]$,
      or equivalently, $[[Γ' ⊢ [σ1](iP → iM) ≈ [σ2](iP → iM)]]$.
    \item $[[iN = ∀pas.iM]]$
      We can assume that $[[pas]]$ is disjoint from $[[Γ]]$ and $[[Γ']]$.
      By inversion, $[[Γ ⊢ ∀pas.iM]]$ implies $[[Γ, pas ⊢ iM]]$.
      Notice that $[[Γ' ⊢ σi : Γ]]$ and 
      $[[Γ' ⊢ σ1 ≈ σ2 : Γ]]$
      can be extended to 
      $[[Γ', pas ⊢ σi : Γ, pas]]$
      and 
      $[[Γ', pas ⊢ σ1 ≈ σ2 : Γ, pas]]$
      by \cref{lemma:subst-domain-weakening}.
      Then by the induction hypothesis,
      $[[Γ', pas ⊢ [σ1]iM ≈ [σ2]iM]]$,
      meaning by inversion that
      $[[Γ', pas ⊢ [σ1]iM ≤ [σ2]iM]]$ and 
      $[[Γ', pas ⊢ [σ2]iM ≤ [σ1]iM]]$.

@@
